Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-04-22 paper/main.tex
Annie through section 1

\noindent \textbf{Keywords:} surrogate, computer experiment, uncertainty quantification,
principal components analysis, Mira-Titan, Cosmic Emu (???)

@@

Computer simulation experiments are invaluable tools in the study of cosmology.
Experiments which simulate the expansion of the universe are growing
in prevalance and complexity \citep[e.g.,][]{lawrence2010coyote,moran2023mira}
\textcolor{red}{(more cites here?)}.  
We are particularly motivated by a simulation
of the matter power spectrum, which describes the distribution of matter as a 
function of spatial scale. 
%It is often represented as a function of wavenumber $k$ (units Mpc$^{-1}$), which is 
%inversely related to spatial scale. 
On large scales, power spectrum behave according to linear perturbation 
theory \citep{pietroni2008flowing, lesgourgues2009non}.  On smaller scales, nonlinear 
dynamics require the use of computationally intensive simulations, yielding
correlated functional data.
%the Coyote Universe \citep{lawrence2010coyote} and the Mira-Titan Universe \citep{moran2023mira} 
%are two simulation suites dedicated to this effort.  
From the highest level, our aim is to leverage such simulation data to 
first estimate and then predict underlying power matter spectra for various 
cosmological parameterizations.

We are particularly motivated by the Mira-Titan simulation suite \citep{moran2023mira}
which defines ``cosmologies'' according to eight cosmological parameters (more on this
in Section \ref{sec:data}).  For a particular cosmology, the simulation of the power matter spectra 
is rather complex.  Although we presume there exists some true matter spectra as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulation outputs 
a convoluted patchwork of potential spectra, which vary in fidelity and which are trusted
only over particular wavenumber ranges.  Our first objective is to synthesize expert knowledge
and all of this simulation information into an effective estimate of the true underlying
matter spectrum.

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_data.png}
    \caption{Left: The perturbation theory, low resolution runs, and high resolution run 
             for the first training cosmology. The weighted average of these is shown as a dashed line. 
             Dotted lines at the bottom indicate where each data type is deemed approximately unbiased. 
             Right: The same image, but restricted to wavenumber ($k$) values where the low resolution 
             runs are approximately unbiased.}
    \label{fig:plot_data}
\end{figure}

There are several challenges to estimating the underlying spectra.  Our model
must: handle multiple correlated functional observations, offer enough 
flexibility to accommodate the nonstationarity present in power matter spectra 
(which vary in smoothenss across wavenumbers), provide effective uncertainty
quantification (UQ), and allow for the incorporation of expert knowledge 
regarding the wavenumbers over which various outputs are trusted.  To this end, 
we propose a Bayesian hierarchical model which treats each of functional simulation 
output as a realization of a deep Gaussian process \citep[DGP;][]{damianou2013deep} 
centered on the true underlying spectra.  The hierarchical structure naturally
accommodates the correlated observations.  The depth of the GP offers nonstationary
flexibility.  The Bayesian framework facilitates UQ and the incorporation of prior
knowledge.  While Bayesian DGPs have been previously deployed for computer experiments 
with scalar responses \citep[e.g.,][]{sauer2023active,sauer2023vecchia,ming2023deep}, 
they have yet to be extended to correlated functional outputs. 
We will demonstrate profiency on synthetic examples and a low fidelity
cosmological simulation, before deploying our model on the motivating 
Mira-Titan simulation.

Our second objective is to leverage simulation data from a limited set of
cosmologies in order to predict power matter spectra for unobserved cosmologies.
Due to the computationally expensive nature of the Mira-Titan simulation suite
\textcolor{red}{(do we know how long it takes to run?)},
it is infeasible to evaluate the simulation for every possible eight-dimensional
cosmological configuration which may be of interest.  Instead, we desire
a statistical ``emulator'' or ``surrogate model'' 
\citep{santner2003design,gramacy2020surrogates} which will provide quick and effective
predictions of power matter spectra for any cosmological parameterization.
\citet{moran2023mira} provide a state-of-the-art emulator for the Mira-Titan
simulation suite, which was trained on 117 simulations, and is termed 
the ``Cosmic Emu''.\footnote{\url{https://github.com/lanl/CosmicEmu}}
We leverage the same training data, in conjunction with our Bayesian hierarchical
DGP model and basis function representations, to train a Gaussian process surrogate 
on the principal component weights in order to predict spectra for unobserved
cosmologies \citep{higdon2008computer, higdon2010estcosmo}. 
Our method compares favorably to Cosmic Emu on held-out cosmologies.

For a comprehensive assessment of our Bayesian hierarchical DGP model, we 
further study its performance in a simulation study with synthetic data mimicing 
the Mira-Titan data. Additionally, we also train our model with data output from 
the Code for Anisotropies in the Microwave Background \citep[CAMB,][]{lewis2011CAMB};
this model output exhibits similar behavior to Mira-Titan as well, but has the benefit 
of including a ``true" power spectrum corresponding toeach cosmology.

The remainder of the paper is organized as follows.  Section \ref{sec:data} 
introduces our motivating application, the Mira-Titan simulation suite.  
Section \ref{sec:hm_fit} describes our hierarchical Bayesian DGP model and 
validates its performance on synthetic exercises and the CAMB simulations.  
Leveraging this trained model, Section \ref{sec:pred} details 
our procedure for predicting at unobserved cosmologies, benchmarking against
state-of-the-art competitors on the CAMB and Mira-Titan simulations. 
Section \ref{sec:disc} concludes with a discussion of our contributions and avenues for 
future research.  We provide reproducible code and an R package for our Bayesian DGP model
in a public git repository.\footnote{\url{https://github.com/stevewalsh124/dgp.hm}}
\textcolor{red}{Please note: this repository is currently private until we 
hear back regarding permissions for sharing the Mira-Titan data alongside our code.}

>>>>

\noindent \textbf{Keywords:} computer experiment, Cosmic Emu, 
principal components analysis, Mira-Titan, surrogate, uncertainty quantification

@@

Computer simulation experiments are invaluable tools in the study of cosmology.
Experiments which simulate the expansion of the universe are growing
in prevalance and complexity \citep[e.g.,][]{lawrence2010coyote,moran2023mira}
\textcolor{red}{(more cites here? possibly those commented out above?)}.  
We are particularly motivated by a simulation
of the matter power spectrum, which describes the distribution of matter as a 
function of spatial scale. 
%It is often represented as a function of wavenumber $k$ (units Mpc$^{-1}$), which is 
%inversely related to spatial scale. 
On large scales, power spectrum behave according to linear perturbation 
theory \citep{pietroni2008flowing, lesgourgues2009non}.  On smaller scales, nonlinear 
dynamics require the use of computationally intensive simulations, yielding
correlated functional data.
%the Coyote Universe \citep{lawrence2010coyote} and the Mira-Titan Universe \citep{moran2023mira} 
%are two simulation suites dedicated to this effort.  
Our overarching goal is to leverage such simulation data to 
first estimate and then predict underlying power matter spectra for various 
cosmological parameterizations.

The Mira-Titan simulation suite \citep{moran2023mira}
defines ``cosmologies'' according to eight cosmological parameters (more on this
in Section \ref{sec:data}).  For a particular cosmology, the simulation of the power matter spectra 
is rather complex.  Although we presume there exists some true matter spectra as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulation outputs 
a convoluted patchwork of potential spectra, which vary in fidelity and which are trusted
only over particular wavenumber ranges.  Our first objective is to synthesize expert knowledge
and all available simulation information into an effective estimate of the true underlying
matter spectrum.  \textcolor{red}{(In this paragraph, we need to narrate Figure 1.  The
figure provides a useful visual, but it needs to be talked through.  Comment on how
the weighted average takes into account the appropriate ranges.  Explain why the weighted
average on its own is not an effective estimate.  Motivate why we need a more complex model.
This is also a good place to introduce wavenumber as $k$.  Is there a particular reason
we plot it on the log scale?)}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth, trim=0 13 0 45, clip=TRUE]{plot_data.png}
    \caption{{\it Left:} The perturbation theory (solid green), low resolution runs (solid grey), 
    and high resolution run  (solid pink) for a particular cosmology. 
    The weighted average of these is shown in dashed orange. Dotted lines at the bottom 
    indicate where each data type is deemed approximately unbiased. 
    {\it Right:} Same, but restricted to wavenumber ($k$) values where the low resolution 
    runs are approximately unbiased.
    \textcolor{red}{(Either make one legend for both graphs or ditch the legend 
    altogether and narrate the colors/line types in the caption.  If legend stays, "ind" needs
    to go.  Pink line is too hard to see.  Make gray lines thinner and pink thicker?
    Since later we will use a centered version, we could add a 3rd panel here that shows the
    same data centered, so the reader can understand the connection to the plots later.)}}
    \label{fig:plot_data}
\end{figure}

There are several challenges to estimating the underlying spectra.  Our model
must: handle multiple correlated functional observations (like the low resolution
runs shown in Figure \ref{fig:plot_data}), offer enough 
flexibility to accommodate the nonstationarity present in power matter spectra 
(which vary in smoothenss across wavenumbers), provide effective uncertainty
quantification (UQ), and allow for the incorporation of expert knowledge 
regarding the wavenumbers over which various outputs are trusted.  To this end, 
we propose a Bayesian hierarchical model which treats functional simulation 
outputs as realizations of a deep Gaussian process \citep[DGP;][]{damianou2013deep} 
centered on the true underlying spectra.  The hierarchical structure naturally
accommodates the correlated observations.  The depth of the GP offers nonstationary
flexibility.  The Bayesian framework facilitates UQ and the incorporation of prior
knowledge.  While Bayesian DGPs have been previously deployed for computer experiments 
with scalar responses \citep[e.g.,][]{sauer2023active,sauer2023vecchia,ming2023deep}, 
they have yet to be extended to correlated functional outputs. 
We will demonstrate profiency on synthetic examples and a low fidelity
cosmological simulation, before deploying our model on the motivating 
Mira-Titan simulation.

Our second objective is to leverage simulation data from a limited set of
cosmologies in order to predict power matter spectra for unobserved cosmologies.
Due to the computationally expensive nature of the Mira-Titan simulation suite
\textcolor{red}{(do we know how long it takes to run?)},
it is infeasible to evaluate the simulation for every possible eight-dimensional
cosmological configuration which may be of interest.  Instead, we desire
a statistical ``emulator'' or ``surrogate model'' 
\citep{santner2003design,gramacy2020surrogates} which will provide quick and effective
predictions of power matter spectra for any cosmological parameterization.
\citet{moran2023mira} provide a state-of-the-art emulator for the Mira-Titan
simulation suite, which was trained on 117 simulations, and is termed 
the ``Cosmic Emu.''\footnote{\url{https://github.com/lanl/CosmicEmu}}
We leverage the same training data, in conjunction with our Bayesian hierarchical
DGP model and basis function representations, to train a Gaussian process surrogate 
on principal component weights in order to predict spectra for unobserved
cosmologies \citep{higdon2008computer, higdon2010estcosmo}. 
Our method compares favorably to Cosmic Emu on held-out cosmologies.

For a comprehensive assessment of our Bayesian hierarchical DGP, we benchmark
its performance in two additional settings: with synthetic data mimicing 
the Mira-Titan data and with real-world data from 
the Code for Anisotropies in the Microwave Background \citep[CAMB,][]{lewis2011CAMB}.
The CAMB model also exhibits similar behavior to Mira-Titan, but crucially
provides a ``true" power spectrum corresponding to each cosmology.

The remainder of the paper is organized as follows.  Section \ref{sec:data} 
introduces our motivating application, the Mira-Titan simulation suite.  
Section \ref{sec:hm_fit} describes our hierarchical Bayesian DGP and 
validates its performance on synthetic exercises and the CAMB simulations.  
Leveraging this trained model, Section \ref{sec:pred} details 
our procedure for predicting at unobserved cosmologies, benchmarking against
state-of-the-art competitors on the CAMB and Mira-Titan simulations. 
Section \ref{sec:disc} concludes with a discussion of our contributions and avenues for 
future research.  We provide reproducible code and an {\sf R} package for our 
Bayesian DGP model in a public git repository.\footnote{\url{https://github.com/stevewalsh124/dgp.hm}}
\textcolor{red}{Please note: this repository is currently private until we 
hear back regarding permissions for sharing the Mira-Titan data alongside our code.}
